Periodograms for Multiband Astronomical Time Series

### 2015-08-26 writeup/paper.tex
remove new/old markers

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \sect{brief_overview}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency \old{and basis functions at different frequencies are not orthogonal. 
As a result, the periodogram has many local maxima.}
\new{and so the likelihood surface in frequency is non-convex.
This non-convexity is readily apparent in the many local maxima of the typical periodogram, which makes it difficult to find the maximum via standard numerical optimization routines.}
Thus in practice the global 
maximum of the periodogram is often found by a brute-force grid search \citep[for details see, e.g.][]{ICVG2014}.

@@

As an example of one of these generalizations, we'll consider 
what has variously been called the {\it Date-compensated Discrete Fourier Transform} \citep{Ferraz-Mello81}, the {\it floating-mean periodogram} \citep{Cumming99}, and the {\it generalized Lomb-Scargle method} \citep{Zechmeister09}. Here we use the term {\it floating-mean periodogram}. This method adjusts the classic normalized periodogram by fitting the mean of the model alongside the amplitudes:
\begin{equation}
  y(t~|~\omega, \theta) = \theta_0 + \theta_1\sin\omega t + \theta_2\cos\omega t
\end{equation}
The periodogram derived from this model can be more accurate than the standard \new{pre-centered} periodogram for certain observing cadences and selection functions, \new{and espeically when searching for long-period varaibility or working with very few samples \citep{Cumming99}.} \citet{Zechmeister09} detail the required modifications to the orthogonalized harmonic formalism of \eq{LombScargle} to allow the mean to float in the model. In the matrix formalism, the modification is much more straightforward: all that is required is to add a column of ones to the $X_\omega$ matrix before computing the power via \eq{LombScargle2}. This column of ones corresponds to a third entry in the parameter vector $\theta$, and acts as a uniform constant offset for all data points.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig03.pdf}
  \caption{
    The model fits and periodograms for several truncated Fourier models.
    The data are the same as those in \fig{basic_example}. Note that
    \new{in addition to the previously-seen 0.38-day alias,} the
    higher-order models will generally show periodogram peaks at multiples
    of the true fundamental frequency $P_0$: this is because for integer $n$
    less than the number of Fourier terms in the model, $P_0$ is a higher
    harmonic of the model at $P=nP_0$. Additionally, the increased degrees of
    freedom in the higher-order models let them fit better at any frequency,
    which drives up the ``background'' level in the periodogram.
  }
  \figlabel{multiterm_example}
\end{figure*}

@@

\new{One might hope that when adding terms, the correct-period model would show more of an improvement than the incorrect-period model (and thus the periodogram maximum would become more pronounced in comparison to the background), but this does not generally hold.
Consider that in the extreme limit in which the number model parameters is equal to the number of data points, the model has enough flexibility to fit the data perfectly at {\it every} frequency, and the resulting periodogram would be everywhere unity!
This can only be the case if, on average, addition of terms preferentially boosts the background level.}

@@

The previous sections raise the question: how complicated a model should we use? We have seen that as we add more terms to the fit, the model will more closely describe the observed data.
\old{In the extreme case, when the number of model parameters equals the number of data points, the model can fit the data {\it exactly} regardless of frequency and the periodogram will be everywhere unity (though in most cases, numerical inaccuracies prevent a truly perfect fit).}
For very high-order models, however, such a close fit {\it over-fits} the data: that is, the fit is more responsive to statistical noise in the observations than to the underlying signal.
This can be addressed by explicitly truncating the series at some number of terms, but we can also use a {\it regularization} term to mathematically enforce model simplicity.

@@

\section{A Multiple-Band Model}
\sectlabel{multiband}
In this section we will combine the ideas of the previous sections to construct the {\it multiband periodogram} which flexibly accounts for heterogeneous sources of data for a single object.
\new{To start with, we might consider one of two na{\"i}ve approaches to the multi-band problem:

@@

Second, we might treat each band entirely independently and compute a standard Lomb-Scargle periodogram on each, and use the additivity of $\chi^2$ along with \eq{chi2PN} to construct a multiband periodogram. This amounts to the assumption that the bands have completely independent phases and amplitudes, and has too many free parameters to be useful in most cases of interest. In other words, the model is too complex and over-fits the data (see \sect{relationship} for further discussion).}

To compute \old{such} a periodogram \new{which strikes a balance between these two extremes}, we will take advantage of the easy extensibility of the matrix formalism which led to our generalizations above.
The multiband model presented here contains the following features:
\begin{enumerate}
\item An $N_{base}$-term truncated Fourier \new{``base model'' which models the shared variability among all $K$ observed bands}. \old{a latent parameter, which we'll call the ``overall variability''.}
  \item A set of $N_{band}$-term truncated Fourier fits, each of which models the residual within a single band from the \new{shared variability accounted for in the base model.} \old{overall variability}.
\end{enumerate}
The total number of parameters for $K$ bands is then $M_K = (2N_{base} + 1) + K(2N_{band} + 1)$. As a result, for each band $k$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  \eqlabel{multiband_model}
  &y_k(t|\omega,\theta) =
  \theta_0 + \sum_{n=1}^{N_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right] +&\nonumber\\ 
  &\theta^{(k)}_0 + \sum_{n=1}^{N_{band}} \left[\theta^{(k)}_{2n - 1}\sin(n\omega t) + \theta^{(k)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(k)}$ are determined individually.
\new{Note the potential for confusion: $N_{band}$ here is not the number of observed bands, but the number of Fourier components fit to the residuals in each of the $K$ observed bands.}

@@

\subsection{Relationship of Multiband and Single-band approaches}
\sectlabel{relationship}
\new{
With this formalism in place, we can return briefly to the na{\"i}ve models discussed at the beginning of \sect{multiband}.
The first, which ignores band information, is simply a standard Lomb-Scargle over the heterogeneous data.
The second, in which each band is fit independently, turns out to be equivalent to the $(N_{base},N_{band})=(0,1)$ model defined above.
}
\old{
The multi-phase model with $(N_{base},N_{band})=(0,1)$ turns out to be a particularly special case.
}
Here the base model is a simple global offset which is degenerate with the offsets in each band, so that the design matrix $X_\omega$ can be straightforwardly rearranged as block-diagonal.
A block-diagonal design matrix in a linear model indicates that components of the model are being solved independently: here these independent components amount to the single-band floating-mean model from \sect{floating_mean}, fit independently for each of the $K$ bands.
\old{This particular multiband model can give us insight into the relationship between single-band and multiband approaches.}

@@

\subsection{Potential Improvements to the Multiband Method}
\new{A well-known (though often unrecognized) difficulty of Lomb-Scargle-type periodograms on unevenly-sampled data is that they do not measure the power of the signal in question, but the power of the signal {\it convolved with the observing with the survey window function}.
For regularly-sampled timeseries, this convolution is the source of the perfect aliasing beyond the Nyquist sampling limit; for non-regular sampling, this aliasing generally happens to some degree at {\it all} frequencies!
Because of this, even a signal with a single well-defined period will result in a Lomb-Scargle periodogram with multiple maxima at locations which depend on both the underlying signal and the precise observing window.

The multiband periodogram, as a generalization of Lomb-Scargle, shares this difficulty:}
\old{The fact that the multiband model does not select the top frequency each time points to its weaknesses: first, as a Fourier-like model,} it tends to respond to frequency structure in the window function as well as frequency structure in the data.
This can be viewed as a result of the very model simplicity which causes its success in the case of sparse multiband data: it cannot disentangle bias in the model from bias due to features in the survey window.

>>>>

The most popular parametric method is the Lomb-Scargle periodogram, which is discussed in detail in \sect{brief_overview}.
The Lomb-Scargle periodogram is related to the $\chi^2$ for a least-square fit of a single sinusoid to data
and can treat non-uniformly sampled time series with heteroscedastic measurement uncertainties. 
The underlying model of the Lomb–Scargle periodogram is nonlinear in frequency and so the likelihood surface in frequency is non-convex.
This non-convexity is readily apparent in the many local maxima of the typical periodogram, which makes it difficult to find the maximum via standard numerical optimization routines.
Thus in practice the global 
maximum of the periodogram is often found by a brute-force grid search \citep[for details see, e.g.][]{ICVG2014}.

@@

As an example of one of these generalizations, we'll consider 
what has variously been called the {\it Date-compensated Discrete Fourier Transform} \citep{Ferraz-Mello81}, the {\it floating-mean periodogram} \citep{Cumming99}, and the {\it generalized Lomb-Scargle method} \citep{Zechmeister09}. Here we use the term {\it floating-mean periodogram}. This method adjusts the classic normalized periodogram by fitting the mean of the model alongside the amplitudes:
\begin{equation}
  y(t~|~\omega, \theta) = \theta_0 + \theta_1\sin\omega t + \theta_2\cos\omega t
\end{equation}
The periodogram derived from this model can be more accurate than the standard \pre-centered periodogram for certain observing cadences and selection functions, and espeically when searching for long-period varaibility or working with very few samples \citep{Cumming99}. \citet{Zechmeister09} detail the required modifications to the orthogonalized harmonic formalism of \eq{LombScargle} to allow the mean to float in the model. In the matrix formalism, the modification is much more straightforward: all that is required is to add a column of ones to the $X_\omega$ matrix before computing the power via \eq{LombScargle2}. This column of ones corresponds to a third entry in the parameter vector $\theta$, and acts as a uniform constant offset for all data points.

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{fig03.pdf}
  \caption{
    The model fits and periodograms for several truncated Fourier models.
    The data are the same as those in \fig{basic_example}. Note that
    in addition to the previously-seen 0.38-day alias, the
    higher-order models will generally show periodogram peaks at multiples
    of the true fundamental frequency $P_0$: this is because for integer $n$
    less than the number of Fourier terms in the model, $P_0$ is a higher
    harmonic of the model at $P=nP_0$. Additionally, the increased degrees of
    freedom in the higher-order models let them fit better at any frequency,
    which drives up the ``background'' level in the periodogram.
  }
  \figlabel{multiterm_example}
\end{figure*}

@@

One might hope that when adding terms, the correct-period model would show more of an improvement than the incorrect-period model (and thus the periodogram maximum would become more pronounced in comparison to the background), but this does not generally hold.
Consider that in the extreme limit in which the number model parameters is equal to the number of data points, the model has enough flexibility to fit the data perfectly at {\it every} frequency, and the resulting periodogram would be everywhere unity!
This can only be the case if, on average, addition of terms preferentially boosts the background level.

@@

The previous sections raise the question: how complicated a model should we use? We have seen that as we add more terms to the fit, the model will more closely describe the observed data.
For very high-order models, however, such a close fit {\it over-fits} the data: that is, the fit is more responsive to statistical noise in the observations than to the underlying signal.
This can be addressed by explicitly truncating the series at some number of terms, but we can also use a {\it regularization} term to mathematically enforce model simplicity.

@@

\section{A Multiple-Band Model}
\sectlabel{multiband}
In this section we will combine the ideas of the previous sections to construct the {\it multiband periodogram} which flexibly accounts for heterogeneous sources of data for a single object.
To start with, we might consider one of two na{\"i}ve approaches to the multi-band problem:

@@

Second, we might treat each band entirely independently and compute a standard Lomb-Scargle periodogram on each, and use the additivity of $\chi^2$ along with \eq{chi2PN} to construct a multiband periodogram. This amounts to the assumption that the bands have completely independent phases and amplitudes, and has too many free parameters to be useful in most cases of interest. In other words, the model is too complex and over-fits the data (see \sect{relationship} for further discussion).

To compute a periodogram which strikes a balance between these two extremes, we will take advantage of the easy extensibility of the matrix formalism which led to our generalizations above.
The multiband model presented here contains the following features:
\begin{enumerate}
\item An $N_{base}$-term truncated Fourier ``base model'' which models the shared variability among all $K$ observed bands. 
  \item A set of $N_{band}$-term truncated Fourier fits, each of which models the residual within a single band from the shared variability accounted for in the base model.
\end{enumerate}
The total number of parameters for $K$ bands is then $M_K = (2N_{base} + 1) + K(2N_{band} + 1)$. As a result, for each band $k$ we have the following model of the observed magnitudes:
\begin{eqnarray}
  \eqlabel{multiband_model}
  &y_k(t|\omega,\theta) =
  \theta_0 + \sum_{n=1}^{N_{base}} \left[\theta_{2n - 1}\sin(n\omega t) + \theta_{2n}\cos(n\omega t)\right] +&\nonumber\\ 
  &\theta^{(k)}_0 + \sum_{n=1}^{N_{band}} \left[\theta^{(k)}_{2n - 1}\sin(n\omega t) + \theta^{(k)}_{2n}\cos(n\omega t)\right].&
\end{eqnarray}
The important feature of this model is that {\it all bands} share the same base parameters $\theta$, while their offsets $\theta^{(k)}$ are determined individually.
Note the potential for confusion: $N_{band}$ here is not the number of observed bands, but the number of Fourier components fit to the residuals in each of the $K$ observed bands.

@@

\subsection{Relationship of Multiband and Single-band approaches}
\sectlabel{relationship}
With this formalism in place, we can return briefly to the na{\"i}ve models discussed at the beginning of \sect{multiband}.
The first, which ignores band information, is simply a standard Lomb-Scargle over the heterogeneous data.
The second, in which each band is fit independently, turns out to be equivalent to the $(N_{base},N_{band})=(0,1)$ model defined above.
Here the base model is a simple global offset which is degenerate with the offsets in each band, so that the design matrix $X_\omega$ can be straightforwardly rearranged as block-diagonal.
A block-diagonal design matrix in a linear model indicates that components of the model are being solved independently: here these independent components amount to the single-band floating-mean model from \sect{floating_mean}, fit independently for each of the $K$ bands.

@@

\subsection{Potential Improvements to the Multiband Method}
A well-known (though often unrecognized) difficulty of Lomb-Scargle-type periodograms on unevenly-sampled data is that they do not measure the power of the signal in question, but the power of the signal {\it convolved with the observing with the survey window function}.
For regularly-sampled timeseries, this convolution is the source of the perfect aliasing beyond the Nyquist sampling limit; for non-regular sampling, this aliasing generally happens to some degree at {\it all} frequencies!
Because of this, even a signal with a single well-defined period will result in a Lomb-Scargle periodogram with multiple maxima at locations which depend on both the underlying signal and the precise observing window.

The multiband periodogram, as a generalization of Lomb-Scargle, shares this difficulty: it tends to respond to frequency structure in the window function as well as frequency structure in the data.
This can be viewed as a result of the very model simplicity which causes its success in the case of sparse multiband data: it cannot disentangle bias in the model from bias due to features in the survey window.
